### 2015-06-18 doc/thesis/system.tex
Back porting abstract syntax from paper to thesis

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rlr}
s ::= & & \textsc{statements:}\\
& \;\; \mathbf{skip} & \textit{skip}\\
& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
& | \; \overline{l} = el & \textit{multiple assignment}\\
& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
& | \; \mathbf{return} \; el & \textit{return} \\
& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
e ::= & & \textsc{expressions:}\\
& \;\; \mathbf{nil} & \textit{nil}\\
& | \; k & \textit{other literals}\\
& | \; id & \textit{variable access}\\
& | \; e_{1}[e_{2}] & \textit{table access}\\
& | \; {<}T{>} \; id & \textit{type coercion}\\
& | \; f & \textit{function declaration}\\
& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
& | \; e_{1} + e_{2} \;
| \; e_{1} \; {..} \; e_{2} \;
| \; e_{1} == e_{2} \;
| \; e_{1} < e_{2} & \textit{binary operations}\\
& | \; e_{1} \;\&\; e_{2} \;
| \; e_{1} \; \mathbf{and} \; e_{2} \;
| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
& | \; \mathbf{not} \; e \;
| \; \# \; e & \textit{unary operations} \\
& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
l ::= & & \textsc{left-hand values:}\\
& \;\; id_{l} & \textit{variable assignment}\\
& | \; e_{1}[e_{2}] & \textit{table assignment}\\
& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
k ::= & & \textsc{literal constants:}\\
& \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} & \\
el ::= & & \textsc{expression lists:}\\
& \;\; \overline{e} \; | \;
\overline{e}, me & \\
me ::= & & \textsc{multiple results:}\\
& \;\; a & \textit{application}\\
& | \; {...} & \textit{vararg expression}\\
a ::= & & \textsc{applications:}\\
& \;\; e(el) & \textit{function application}\\
& | \; e{:}n(el) & \textit{method application}\\
f ::= & & \textsc{function declarations:}\\
& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
pl ::= & & \textsc{parameter lists:}\\
& \;\; \overline{id{:}T} \; | \;
\overline{id{:}T},{...}{:}T & \\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$k$ are literal constants, $el$ are expression lists,
$me$ are expressions with multiple results, $a$ are function and method applications,
$f$ are function declarations, $pl$ are parameter lists,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$.

@@

We also include two kinds of type coercions in our core language:
the left-hand value $id[k] \; {<}T{>}$ and the expression ${<}T{>} \;id$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $id$ when variable names appear as expressions and $id_{l}$ when
variable names appear as left-hand values.

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
We use two different environments to handle this feature.
The first environment is the type environment $\env$ that maps
variables to first-level types.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
The second environment is the projection environment $\penv$ that
maps projection variables to second-level types.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$
with the projection variable $x$ that maps to type $S$.
In Section \ref{sec:fap} we will show how our type system uses the
projection environment $\penv$ for handling projection types,
and also for projecting unions of second-level types into
unions of first-level types.

@@

In Section \ref{sec:refinement} we will show that table refinement can
change the type environment while typing an expression or a left-hand value.
Thus, the rules that type check lists of expressions and lists of
left-hand values use a partial auxiliary function \emph{merge} to collect
all environment changes in a new environment $\env_{m}$, if there are no conflicts.
In Section \ref{sec:refinement} we will also show that we can only change the
environment to add new table fields in a table type, and we cannot change the
type of a variable or a table field which is already present in a table type.

@@

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that expresses this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

@@

Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $T{*}$.
For instance, the rules \textsc{T-EXPLIST2} and \textsc{T-EXPLIST3} handle
the case where an expression list ends with an expression that may produce
multiple values.
The former rule includes the type $\Nil{*}$, but the latter rule does not:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m} \times \Void, \env_{n+2}\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m} \times \Nil{*}, \env_{f}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{1}, \penv \vdash me:T_{n+1} \times ... \times T_{n+m}{*}, \env_{n+2}\\
       \env_{f} = merge(\env_{1}, ..., \env_{n+2}) \;\;\;
       n = |\;\overline{e}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \overline{e},me:T_{1} \times ... \times T_{n+m}{*}, \env_{f}}
\end{array}
\end{array}
\]

Rules for function applications are similar to the rule for multiple assignment.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
       S_{3} \lesssim S_{1}}
      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
that produce no value or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
\end{array}
\end{array}
\]

We can define \emph{first} inductively as follows:
\begin{align*}
first(\Void) & = \Nil\\
first(T{*}) & = T \cup \Nil\\
first(T \times P) & = T\\
first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
\end{align*}

@@

Our type system also catches arity mismatch.
To do that, we end the input type of a function with type $\Void$
instead of $\Value{*}$.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]

@@

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls without optional parameters.
For instance, let us assume that $f$ has type
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$.
The function call
\[
f(1)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]
and the function call
\[
f(1,2,3)
\]
also does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]

>>>>

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rlr}
s ::= & & \textsc{statements:}\\
& \;\; \mathbf{skip} & \textit{skip}\\
& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
& | \; \overline{l} = el & \textit{multiple assignment}\\
& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
& | \; \mathbf{rec} \; id{:}F = e \; \mathbf{in} \; s & \textit{recursive declaration} \\
& | \; \mathbf{return} \; el & \textit{return} \\
& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
e ::= & & \textsc{expressions:}\\
& \;\; \mathbf{nil} & \textit{nil}\\
& | \; k & \textit{other literals}\\
& | \; id & \textit{variable access}\\
& | \; e_{1}[e_{2}] & \textit{table access}\\
& | \; {<}F{>} \; id & \textit{type coercion}\\
& | \; f & \textit{function declaration}\\
& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
& | \; e_{1} + e_{2} \;
| \; e_{1} \; {..} \; e_{2} \;
| \; e_{1} == e_{2} \;
| \; e_{1} < e_{2} & \textit{binary operations}\\
& | \; e_{1} \;\&\; e_{2} \;
| \; e_{1} \; \mathbf{and} \; e_{2} \;
| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
& | \; \mathbf{not} \; e \;
| \; \# \; e & \textit{unary operations} \\
& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
l ::= & & \textsc{left-hand values:}\\
& \;\; id_{l} & \textit{variable assignment}\\
& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
k ::= & & \textsc{literal constants:}\\
& \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} & \\
el ::= & & \textsc{expression lists:}\\
& \;\; \overline{e} \; | \;
\overline{e}, me & \\
me ::= & & \textsc{multiple results:}\\
& \;\; a & \textit{application}\\
& | \; {...} & \textit{vararg expression}\\
a ::= & & \textsc{applications:}\\
& \;\; e(el) & \textit{function application}\\
& | \; e{:}n(el) & \textit{method application}\\
f ::= & & \textsc{function declarations:}\\
& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
pl ::= & & \textsc{parameter lists:}\\
& \;\; \overline{id{:}F} \; | \;
\overline{id{:}F},{...}{:}F & \\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$k$ are literal constants, $el$ are expression lists,
$me$ are expressions with multiple results, $a$ are function and method applications,
$f$ are function declarations, $pl$ are parameter lists,
$id$ are variable names, $F$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}F}$ denotes the list $id_{1}{:}F_{1}, ..., id_{n}{:}F_{n}$.

@@

We also include two kinds of type coercions in our core language:
the left-hand value $id[e] \; {<}V{>}$ and the expression ${<}F{>} \;id$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $id$ when variable names appear as expressions and $id_{l}$ when
variable names appear as left-hand values.

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
We use two different environments to handle this feature.
The first environment is the type environment $\env$ that maps
variables to expression types, as the type of an expression can
be a first-level type, a filter type, or a projection type.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
The second environment is the projection environment $\penv$ that
maps projection variables to second-level types.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$
with the projection variable $x$ that maps to type $S$.
In Section \ref{sec:fap} we will show how our type system uses the
projection environment $\penv$ for handling projection types,
and also for projecting unions of second-level types into
unions of first-level types.

@@

In Section \ref{sec:refinement} we will show that table refinement can
change the type environment while typing an expression or a left-hand value.
Thus, the rules that type check lists of expressions and lists of
left-hand values use a partial auxiliary function \emph{merge} to collect
all environment changes in a new environment $\env_{f}$, if there are no conflicts.
In Section \ref{sec:refinement} we will also show that we can only change the
environment to add new table fields in a table type, and we cannot change the
type of a variable or a table field which is already present in a table type.

@@

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that expresses this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:E_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:E_{2}, \env_{3} \;\;\;
       E_{1} \lesssim E_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

@@

Rules for function applications are similar to the rule for multiple assignment.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:E_{1} \rightarrow E_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:E_{3}, \env_{3} \;\;\;
       E_{3} \lesssim E_{1}}
      {\env_{1}, \penv \vdash e(el):E_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
that produce no value or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):E, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(E), \env_{2}}
\end{array}
\end{array}
\]

We can define \emph{first} inductively as follows:
\begin{align*}
first(F{*}) & = F \cup \Nil\\
first(F \times P) & = F\\
first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
\end{align*}

@@

Our type system also catches arity mismatch.
To do that, we end the input type of a function with type $\Nil{*}$
instead of $\Value{*}$.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Nil{*} \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Nil{*}
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Nil{*}
\]

@@

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls without optional parameters.
For instance, let us assume that $f$ has type
$\Integer \times \Integer \times \Nil{*} \rightarrow \Integer \times \Nil{*}$.
The function call
\[
f(1)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Nil{*}
\]
and the function call
\[
f(1,2,3)
\]
also does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Nil{*}
\]
